The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, ``Parallel Algorithms'' lecture, 2012 winter term at TU Wien.}

### 2014-03-12 article/article.tex
fix punctuation after equations
reviewing first sections once more

\title{The Power of Two Random Choices\\ 
\large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}}
\author{Martin Kalany, 0825673}

\maketitle
\begin{abstract}
Balls-into-bins is a classical allocation problem, where $m$ balls have to be distributed among $n$ bins as evenly as possible. A well studied approach is to select a bin uniformly and independently at random for each ball. Another strategy is to select $d\geq 2$ bins for each ball and put the ball into the bin with the least load. Somewhat surprisingly, the maximum load achieved by such a multiple-choice scheme is significantly smaller, even for $d=2$. In this paper, we will present two different multiple-choice allocation strategies for the balls-into-bins problem and provide a detailed analysis of the upper bounds on the maximum load that they achieve. We will use a technique called witness trees, which has proven useful in a wide variety of balls-into-bins problems.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Suppose that $m$ balls are to be thrown into $n$ bins, according to some placement strategy. We are interested in the number of balls in the fullest bin, also called the \emph{maximum load}. Equivalently, one can look at the \emph{additive gap} of the allocation, i.e., the difference of the number of balls in the fullest bin and the average amount of balls in the bins. These balls-into-bins games (BiB) have been studied extensively (see, e.g.,~\cite{JK77}). In the context of Computer Science, the BiB model is useful to study, e.g., strategies for distributing tasks (balls) to processors (bins) as evenly as possible without having to query all the processors for the number of tasks currently assigned to them.

While most of this survey and the analyses focuses on the lightly loaded case where $m=n$, we will give results for the case $m > n$ as well. We will denote the maximum load by $\load$ and the additive gap by $\delta$:
\begin{align*}
\delta = \load - \frac{m}{n}.
\end{align*}

\subsection{Single-choice BiB}
\label{sec:single-choice}
The \emph{Single-choice} strategy is a simple allocation scheme for the BiB problem and chooses one bin uniformly and independently at random for each ball. Using this strategy, the maximum load of the fullest bin will be
\begin{comment}
\footnote{A more accurate result exists: The maximum load is $\load = \Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$~\cite{G91}.}
\end{comment}
\begin{align*}
\load = \frac{\ln n}{\ln \ln n}  \left(1 + o\left(1\right)\right)
\end{align*}
for the case $m=n$, with high probability\footnote{We say that an event $\mathcal E_n$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E_n \right] \geq 1 - n^{-\alpha}$ for some arbitrary constant $\alpha > 0$.}~\cite{RS98}. This implies (see Theorem \ref{theorem:asymptotic} in Appendix \ref{sec:asymptotic})
\begin{align*}
\load\in \Theta\left(\frac{\ln n}{\ln \ln n}\right).
\end{align*}
Note that, since the average load is 1, $\delta \in \Theta\left( L_{max} \right)$ in this case. For the heavily loaded case, w.h.p., the maximum gap is ~\cite{RS98}
\begin{align*}
\delta = \Theta\left(\sqrt{\frac{m  \ln n}{n}}\right).
\end{align*}

\subsection{Multiple-choice BiB}
Consider a variation of the  single-choice BiB strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the least full bin. If there are multiple bins with the same minimal load, chose one among them uniformly and independently at random. This simple multiple-choice scheme, called the \emph{Greedy} algorithm, achieves a gap
\begin{align*}
\delta = \frac{\ln \ln n}{\ln d} + O(1),
\end{align*}
w.h.p., for all $m$~\cite{ABKU99, BCSV06}, i.e., this additional choice leads to an exponential reduction in the maximum load. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influential paper by Azar et al.~\cite{ABKU99}, who were the first ones to explicitly prove this phenomenon\footnote{For a brief history of the proof, which is quite interesting by itself, refer to Appendix \ref{sec:historyOfGreedy}.}, suggests that this \emph{two-choice paradigm} is not just an artifact of the simple BiB model, but a general principle applicable to a variety of problems.

Consider the slightly more involved \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ (To simplify the presentation, we will assume that $d$ divides $n$ evenly in the remainder of this survey). We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with same minimal load. Somewhat surprisingly, V\"ocking~\cite{VOC03} showed that such a strategy employing asymmetric tie-breaking in combination with non-uniform bin selection, reduces the bound for the gap to\
\begin{align*}
\delta = \frac{\ln\ln n}{d  \ln \Phi_d} + O(1),
\end{align*}
w.h.p., for all $m$. (The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.) Not only is this a significant reduction of the asymptotic bound, V\"ocking was able to show that the Always-go-left strategy is optimal up to additive constants~\cite{VOC03}.

\emph{Outline:}
In Section \ref{sec:classification}, we give a basic classification of BiB games to provide an overview of the research, followed by some applications in Section \ref{sec:applications}. Section \ref{sec:model} gives a detailed definition of the model we use, which can be classified as a sequential, finite and dynamic process with random deletions. In Section \ref{sec:multiple-choice} we provide a brief definition of three different classes of random placement strategies and present two placement algorithms outlined in the introduction: The Greedy algorithm (Section \ref{sec:greedy}) and the Always-go-left algorithm (Section \ref{sec:AlwaysGoLeft}). Section \ref{sec:analysis} contains a detailed analysis of both strategies for the case $m=n$.

\section{Classification}
\label{sec:classification}
Several different variations of this problem exist, which may be classified as follows:
\begin{compactitem}
\item In a \emph{sequential} process, the balls arrive one after another and ball is dealt with only after the previous one has been placed. Conversely, in a \emph{parallel} process, balls arrive in sets and the balls in one set are dealt with simultaneously~\cite{ABS98}. 
\item In a \emph{static} problem, the balls are placed into the bins and no balls are ever removed, whereas in a \emph{dynamic} problem balls may also be removed from the system. In the random deletions model, for each newly inserted ball another one is chosen uniformly and independently at random and removed from the system~\cite{ABKU99, MRS01}. More elaborate schemes exists: An oblivious adversary might specify the sequence of insertions and deletions of balls in advance, without any knowledge of their placement~\cite{CFM+98}. Adler et al.~\cite{ABS98} studied another approach in the context of the parallel BiB problem, where the balls are stored in FIFO-order and at every time step the first ball in each of the bins is deleted. 
\item A \emph{finite} process is analyzed for a given time interval that is known beforehand, whereas for an \emph{infinite} one, placements and deletions are given for an infinite time span~\cite{ABS98}.
\item In the lightly loaded case, $m$ is in the order of $n$, whereas in the heavily loaded case, $m$ is arbitrarily large. W.l.o.g.~we will assume that $m = n$ for the lightly loaded case. 
\end{compactitem}

\section{Applications}
\label{sec:applications}
The two-choice paradigm applied to the BiB model has several useful applications, some of which we will outline here. Hashing (Section \ref{sec:hashing}), online load balancing (Section \ref{sec:loadbalancing}) and the emulation of shared memory machines on distributed memory machines (Section \ref{sec:DMM}) share a close relation to load balancing. Section \ref{sec:circuitrouting} applies the BiB model and the two-choice paradigm to a different context, namely routing circuits in multi-stage interconnection networks with minimal congestion.

\subsection{Hashing}
\label{sec:hashing}
A simple hash table implementation typically uses a single hash function to map elements to entries in a table. If more than one element is mapped to the same entry, all elements mapped to this table entry are stored in a chain, usually implemented as a linked list. The maximum time to search for an element is then proportional to the length of the longest chain in the hash table.

It is easy to see how the BiB model can be used: Assuming a perfect hash function\footnote{I.e., each element is mapped to one of the $n$ table entries uniformly and independently at random}, the elements can be seen as balls to be mapped to table entries (bins), implying a worst case look-up time of $\Theta\left(\frac{\ln n}{\log \ln n}\right)$, w.h.p., where $n$ denotes the number of elements. Applying the two-choice paradigm by using two hash functions and mapping elements to the least full of two table entries, results in a look-up time of $\Theta\left(\ln \ln n\right)$, w.h.p. This approach is easy to parallelize, it does not require re-hashing of data and requires only two hash functions~\cite{ABKU99, MRS01}.

@@

To maximize the throughput of the servers, a uniform distribution of tasks to servers is desirable. While a central dispatcher can easily achieve this, it also poses a bottleneck within a distributed system, since querying $m$ servers for their current load involves significant overhead (e.g., sending a message to each server and waiting for the reply). As with hashing, the BiB model can be applied easily: we need to assign tasks (the balls) to servers (the bins). Assuming tasks are assigned to servers uniformly and independently at random, the two-choice paradigm is directly applicable: If each client samples two random servers for their load and sends its request to the least loaded server, the overhead is comparatively small and the difference in load on the $n$ servers is at most $O\left(\ln \ln n\right)$, w.h.p., \cite{KLM92,RS98, MRS01}.

\subsection{Emulation of shared memory machines on distributed memory machines}
\label{sec:DMM}
The Parallel Random Access Machine model (PRAM) ~\cite{P03} is a high level abstraction of a shared memory machine, in which memory access by any processor to any memory location is assumed to require constant time. It is extremely useful for the study of parallel algorithms. However, a PRAM is much harder to implement in hardware than a Distributed Memory Machine (DMM). Thus, ways to emulate a PRAM on a DMM have been studied extensively. Such an emulation distributes the processors and memory cells of the PRAM to the processors and memory modules (typically one per processor) of the DMM, with the goal to minimize the slowdown of the emulation, i.e., the time the DMM requires to simulate one step of the PRAM execution. Each memory module can handle only one memory access at a time step and more accesses to the same module will result in memory contention. The two-choice paradigm is useful for managing communication between the processors and the memory modules such that memory contention is minimal~\cite{KLM92}.

\subsection{Low congestion circuit routing}
\label{sec:circuitrouting}
In multi-stage interconnection networks, virtual circuit-switching is used to route communication. For simplicity, we consider the permutation routing problem, where one request originates at each of the $n$ input nodes and one request is destined for each of the $n$ output nodes and we assume a butterfly network as the underlying interconnection network~\cite{CLR09}. The congestion is the maximum number of paths that are routed through a network link and the goal of a circuit routing algorithm is to to allocate paths for all $n$ communication requests such that link congestion is minimized.

In this context, Valiant's paradigm~\cite{V82} states that any such permutation routing problem can be solved by transforming it into two problems: First, find a path for each request to an intermediate destination chosen uniformly and independently at random. Next, find a path from the intermediate to the actual destination. This routing technique is analogous to the single-choice BiB problem where each communication request (or ball) chooses an intermediate destination (a bin) at random.
Extensive research~\cite{CMM+98,MRS01} shows that by applying the two-choice paradigm, the link congestion can be bounded by $\Theta\left(\log \log n\right)$ , w.h.p.

\section{BiB model}
\label{sec:model}
Although several different variants of the BiB problem exist, we will focus on a basic model to highlight the crucial ideas in the placement strategies and their analyses. They are based on a sequential, infinite and dynamic process that uses an oblivious adversary, which provides a sequence $\sigma = \sigma_1 \sigma_2 \dots$, where $\sigma_t = (d_t, a_t)$ represents a request for a ball $d_t$ to be deleted and a different ball $a_t$ to be added to the bins at time step $t$. In our model, the ball $b_d$ associated with some request $\sigma_t$ is deleted at the beginning of time step $t$. Next, the ball $b_a$ has to be placed into one of the $n$ bins without knowledge of the following requests $\sigma_{t+1}$, $\sigma_{t+2}\dots$. This infinite sequence of insertions and deletions is fixed in advance and is thus independent of previously allocated balls and any decisions made by the allocation algorithm. We assume that at any time $t$ at most $m = n$ balls are in the bins.

\section{Multiple-choice BiB}
\label{sec:multiple-choice}
The single-choice BiB strategy (Section \ref{sec:single-choice}) chooses one bin uniformly and independently at random for each ball. A multiple-choice BiB strategy generalizes this idea by  choosing $d \geq 2$ bins and placing the ball into the least full bin. We distinguish placement algorithms depending on how the $d$ locations are sampled~\cite{VOC03}.

\subsection{Classes of random placement strategies}
\label{sec:classesOfPlacement}
Let $B$ be the set of bins $\{1,2,...,n\}$ and $B^{d}$ be the sample space. Placement algorithms are distinguished by the way the $d$ candidate bins are chosen for each ball:
\begin{description}
\item [Uniform and independent.] Choose each of the $d$ bins for a ball uniformly and independently at random from the set of bins $B$. Thus, the probability that any specifc bin $b$ is choosen for the $i$-th balls is
\begin{align*}
\Pr[\mathrm{bin}\;b\;\mathrm{choosen}] = \frac{1}{n}
\end{align*}
\item [Non-uniform and independent.] Choose the $i$-th bin from $B$ independently at random according to some probability distribution: 
\item [Non-uniform and dependent.] Choose the $d$ bins for a ball from the sample space according to some probability probability distribution:
\end{description}

We will discuss two different multiple-choice placement strategies: The Greedy algorithm (Section \ref{sec:greedy}), belonging to class 1, and the Always-go-left scheme (Section \ref{sec:AlwaysGoLeft}), which uses a non-uniform and independent selection of $d$ bins and thus belongs to class 2.

\subsection{The Greedy scheme}
\label{sec:greedy}
Consider the Greedy scheme, a variation of the one-choice BiB strategy: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the least full bin. If there are multiple bins with the same minimal load, chose one among them arbitrarily. Note that the $d$ selected bins are not necessarily distinct.

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, the maximum load is at most
\begin{align*}
\load = \frac{\ln \ln n}{\ln d}+ O(1)
\end{align*}
w.h.p., for all $t$~\cite{ABKU99, BCSV06}.
\end{theorem}

This result is remarkable in several ways:
\begin{compactitem}
\item The apparently small change made for the Greedy algorithm compared to the single-choice BiB scheme results in an exponential decrease of the gap, even for $d=2$. 
\item Each additional choice ($d > 2$) decreases the gap by only a constant factor~\cite{MRS01}.
\end{compactitem}

\subsection{The Always-go-left scheme}
\label{sec:AlwaysGoLeft}
The Greedy strategy presented in Section~\ref{sec:greedy} assumes that $d$ bins are chosen uniformly and independently at random. V\"ocking~\cite{VOC03} was the first to study the impact of non-uniform and dependent selections of the $d$ bins. Surprisingly, a multiple-choice BiB strategy with a non-uniform and independent selection of the $d$ bins, such as the Always-go-left algorithm introduced in~\cite{VOC03}, achieves better~\cite{BCSV06} load balancing than the Greedy scheme.

The Always-go-left algorithm works as follows. Partition the $n$ bins into $d$ groups, each of size $\frac{n}{d}$. We select one from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $b_i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with same minimal load.

\begin{comment}
\begin{theorem}
\label{theorem:agln}
If $n$ balls are placed into $n$ bins using the Always-go-left scheme, the maximum load will be
\begin{align*}
\load \leq \frac{\ln\ln n}{d  \ln \Phi_d} + O(1)
\end{align*}
w.h.p.
\end{theorem}
\end{comment}

\begin{theorem}
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load is at most 
\begin{align*}
\load = \frac{\ln\ln n}{d  \ln \Phi_d} + O(1)
\end{align*}
w.h.p.,~for all $t$~\cite{VOC03}.
\end{theorem}

\section{Analysis}
\label{sec:analysis}
In this section, we will prove Theorems \ref{theorem:greedy} and \ref{theorem:aglm} in detail, using witness trees to provide an upper bound for the event that a bin contains too many balls. Our analysis closely follows the proofs in~\cite{VOC03}.

When the maximum load exceeds some given threshold value we say that a bad event occurs and construct an activated witness tree. Thus, the activation of a witness tree is implied by a bad event. The probability for the existence of such an activated witness tree upper-bounds the probability that the associated bad event occurs.

Throughout this section, ``location'' will denote a specific place in one of the $n$ bins. A location of a ball $b$ refers to a specific bin and to the number of balls already in it (at the time of insertion of ball $b$).

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections~\ref{sec:analysis:greedy} and~\ref{sec:analysis:alg}, respectively. However, we use a simplifying assumption (Assumption~\ref{assumption:independence}) in those two sections to provide a more clear presentation of the proof technique. Section~\ref{sec:analysis:nondistinctBalls} will extend these proofs to remove the simplifying assumption we make initially.

We will use a proof technique called witness trees, which allows for a clean analysis of both problems and is applicable to a wide variety of BiB scenarios. Other proof techniques such as layered induction and a differential equations approach have been used. See, e.g.,~\cite{MRS01} for an overview of different proof techniques for the BiB problem. The proof for arbitrarily large $m$ is more complicated and we refer the interested reader to~\cite{BCSV06, TW13}.

>>>>

\title{The Power of Two Random Choices\\ 
\large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, ``Parallel Algorithms'' lecture, 2012 winter term at TU Wien.}}
\author{Martin Kalany, 0825673}
\maketitle

\begin{abstract}
Balls-into-bins is a classical allocation problem, where $m$ balls have to be distributed among $n$ bins as evenly as possible. A well studied approach is to select a bin uniformly and independently at random for each ball. Another strategy is to select $d\geq 2$ bins for each ball and put the ball into the bin with the least load. Somewhat surprisingly, the maximum load achieved by such a multiple-choice scheme is significantly smaller, even for $d=2$. In this paper, we will present two different multiple-choice allocation strategies for the balls-into-bins problem and provide a detailed analysis of the upper bounds on the maximum load that they achieve. We will use a technique called witness trees, which has proven useful in a wide variety of balls-into-bins problems.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Suppose that $m$ balls are to be thrown into $n$ bins, according to some placement strategy. We are interested in the number of balls in the fullest bin, also called the \emph{maximum load}. Equivalently, one can look at the \emph{additive gap} of the allocation, i.e., the difference of the number of balls in the fullest bin and the average amount of balls in the bins. These balls-into-bins games (BiB) have been studied extensively (see, e.g.,~\cite{JK77}). In the context of computer science, the BiB model is useful to study, e.g., strategies for distributing tasks (balls) to processors (bins) as evenly as possible without having to query all the processors for the number of tasks currently assigned to them.

While most of this survey and the analyses focuses on the lightly loaded case where $m=n$, we will give results for the case $m > n$ as well. We will denote the maximum load by $\load$ and the additive gap by $\delta$:
\begin{align*}
\delta = \load - \frac{m}{n} \quad .
\end{align*}

\subsection{Single-choice BiB}
\label{sec:single-choice}
The \emph{Single-choice} strategy is a simple allocation scheme for the BiB problem and chooses one bin uniformly and independently at random for each ball. Using this strategy, the maximum load of the fullest bin will be
\begin{align*}
\load = \frac{\ln n}{\ln \ln n}  \left(1 + o\left(1\right)\right)
\end{align*}
for the case $m=n$, with high probability\footnote{We say that an event $\mathcal E_n$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E_n \right] \geq 1 - n^{-\alpha}$ for some arbitrary constant $\alpha > 0$.}~\cite{RS98}. This implies (see Theorem \ref{theorem:asymptotic} in Appendix \ref{sec:asymptotic})
\begin{align*}
\load\in \Theta\left(\frac{\ln n}{\ln \ln n}\right)\quad .
\end{align*}
Note that, since the average load is 1, $\delta \in \Theta\left( L_{max} \right)$ in this case. For the heavily loaded case~\cite{RS98}, the maximum gap is, w.h.p.,
\begin{align*}
\delta = \Theta\left(\sqrt{\frac{m  \ln n}{n}}\right)\quad .
\end{align*}

\subsection{Multiple-choice BiB}
Consider a variation of the  single-choice BiB strategy: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the least full bin. If there are multiple bins with the same minimal load, choose one among them arbitrarily. This simple multiple-choice scheme, called the \emph{Greedy} algorithm, achieves a maximum load of
\begin{align*}
\load = \frac{m}{n}+\frac{\ln \ln n}{\ln d} + O(1) \Rightarrow \delta = \frac{\ln \ln n}{\ln d} + O(1)
\end{align*}
w.h.p., for all $m$~\cite{ABKU99, BCSV06}, i.e., this additional choice leads to an exponential reduction of the additional gap. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influential paper by Azar et al.~\cite{ABKU99}, who were the first ones to explicitly prove this phenomenon\footnote{For a brief history of the proof, which is quite interesting by itself, refer to Appendix \ref{sec:historyOfGreedy}.}, suggests that this \emph{two-choice paradigm} is not just an artefact of the simple BiB model, but a general principle applicable to a variety of problems.

Consider the slightly more involved \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ (to simplify the presentation, we will assume that $d$ divides $n$ evenly in the remainder of this survey). We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with the same minimal load. Somewhat surprisingly, V\"ocking~\cite{VOC03} showed that such a strategy employing asymmetric tie-breaking in combination with non-uniform bin selection reduces the bound for the gap to\
\begin{align*}
\delta = \frac{\ln\ln n}{d  \ln \Phi_d} + O(1),
\end{align*}
w.h.p., for all $m$. (The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.) V\"ocking was also able to show that the Always-go-left strategy is optimal up to additive constants~\cite{VOC03}.

\subsubsection*{Outline}
In Section \ref{sec:classification}, we give a basic classification of BiB games to provide an overview of the research, followed by some applications in Section \ref{sec:applications}. Section \ref{sec:model} contains a detailed definition of the model we use, which can be classified as a sequential, finite and dynamic process with random deletions. In Section \ref{sec:multiple-choice} we provide a brief definition of three different classes of random placement strategies and present two placement algorithms outlined in the introduction: Greedy (Section \ref{sec:greedy}) and Always-go-left (Section \ref{sec:AlwaysGoLeft}). Section \ref{sec:analysis} contains a detailed analysis of both strategies for the case $m=n$.

\section{Classification of BiB games}
\label{sec:classification}
Several different variations of this problem exist, which may be classified as follows:
\begin{compactitem}
\item In a \emph{sequential} process, the balls arrive one after another and a ball is dealt with only after the previous one has been placed. Conversely, in a \emph{parallel} process, balls arrive in sets and the balls in one set are dealt with simultaneously~\cite{ABS98}. 
\item In a \emph{static} problem, the balls are placed into the bins and no balls are ever removed, whereas in a \emph{dynamic} problem balls may also be removed from the system. In the random deletions model, for each newly inserted ball another one is chosen uniformly and independently at random and removed from the system~\cite{ABKU99, MRS01}. More elaborate schemes exists: An oblivious adversary might specify a sequence of insertions and deletions of balls in advance, without any knowledge of their placement~\cite{CFM+98}. Adler et al.~\cite{ABS98} studied another approach in the context of the parallel BiB problem, where the balls are stored in FIFO-order and at every time step the first ball in each bin is deleted. 
\item A \emph{finite} process is analysed for a given time interval that is known beforehand, whereas for an \emph{infinite} one, placements and deletions are given for an infinite time span~\cite{ABS98}.
\item In the lightly loaded case, $m$ is in the order of $n$, whereas in the heavily loaded case, $m$ is arbitrarily large. W.l.o.g., we will assume that $m = n$ for the lightly loaded case. 
\end{compactitem}

\section{Applications}
\label{sec:applications}
The two-choice paradigm applied to the BiB model has several useful applications, e.g., hashing (Section \ref{sec:hashing}), online load balancing (Section \ref{sec:loadbalancing}), the emulation of shared memory machines on distributed memory machines (Section \ref{sec:DMM}) and routing circuits in multi-stage interconnection networks with minimal congestion (Section~\ref{sec:circuitrouting}).

\subsection{Hashing}
\label{sec:hashing}
A basic hash table implementation typically uses a single hash function to map elements to entries in a table. If several elements are mapped to the same entry, they are stored in a chain, usually implemented as a linked list. The maximum time to search for an element is then proportional to the length of the longest chain in the hash table.

It is easy to see how the BiB model can be applied: Assuming a perfect hash function\footnote{I.e., each element is mapped to one of the $n$ table entries uniformly and independently at random}, the elements can be seen as balls to be mapped to table entries (bins), thus implying a worst case look-up time of $\Theta\left(\frac{\ln n}{\ln \ln n}\right)$, w.h.p., where $n$ denotes the number of elements. Applying the two-choice paradigm by using two hash functions and mapping elements to the least full of two table entries results in a look-up time of $\Theta\left(\ln \ln n\right)$, w.h.p. This approach is easy to parallelize, it does not require re-hashing of data and uses only two hash functions~\cite{ABKU99, MRS01}.

@@

To maximize the throughput of the servers, a uniform distribution of tasks to servers is desirable. While a central dispatcher can easily achieve this, it also poses a bottleneck within a distributed system, since querying $m$ servers for their current load involves significant overhead (e.g., sending a message to each server and waiting for the reply). The BiB model can be applied easily: we need to assign tasks (the balls) to servers (the bins). Assuming tasks are assigned to servers uniformly and independently at random, the two-choice paradigm is directly applicable: If each client samples two random servers for their load and sends its request to the least loaded server, the overhead is comparatively small and the difference in load on the $n$ servers is $\Theta\left(\ln \ln n\right)$, w.h.p., \cite{KLM92,RS98, MRS01}.

\subsection{Emulation of shared memory machines on distributed memory machines}
\label{sec:DMM}
The Parallel Random Access Machine model (PRAM) ~\cite{P03} is a high level abstraction of a shared memory machine, in which memory access by any processor to any memory location is assumed to require constant time. It is extremely useful for the study of parallel algorithms. However, a PRAM is much harder to implement in hardware than a Distributed Memory Machine (DMM). Thus, ways to emulate a PRAM on a DMM have been studied extensively. Such an emulation distributes the processors and memory cells of the PRAM to the processors and memory modules of the DMM, with the goal to minimize the slowdown of the emulation, i.e., the time the DMM requires to simulate one step of the PRAM execution. Each memory module can handle only one memory access at a time step and more accesses to the same module result in memory contention. The two-choice paradigm is useful for managing communication between the processors and the memory modules such that memory contention is minimized~\cite{KLM92}.

\subsection{Low congestion circuit routing}
\label{sec:circuitrouting}
In multi-stage interconnection networks, virtual circuit-switching is used to route communication. For simplicity, we consider the permutation routing problem, where one request originates at each of the $n$ input nodes and one request is destined for each of the $n$ output nodes. Furthermore, we assume a butterfly network as the underlying interconnection network~\cite{CLR09}. The link congestion is the maximum number of paths that are routed through a network link and the goal of a circuit routing algorithm is to allocate paths for all $n$ communication requests such that the link congestion is minimized.

In this context, Valiant's paradigm~\cite{V82} states that any such permutation routing problem can be solved by transforming it into two sub-problems: First, find a path for each request to an intermediate destination chosen uniformly and independently at random. Next, find a path from the intermediate to the actual destination. This routing technique is analogous to the single-choice BiB problem where each communication request (or ball) chooses an intermediate destination (a bin) at random.
Extensive research~\cite{CMM+98,MRS01} shows that by applying the two-choice paradigm, the link congestion can be bounded by $\Theta\left(\log \log n\right)$ , w.h.p.

\section{BiB model}
\label{sec:model}
Although several different variants of the BiB problem exist, we will focus on a basic model to highlight the crucial ideas in the placement strategies and their analyses. They are based on a sequential, infinite and dynamic process that uses an oblivious adversary, which provides a sequence $\sigma = \sigma_1 \sigma_2 \dots$, where $\sigma_t = (d_t, a_t)$ represents a request for a ball $d_t$ to be deleted and a different ball $a_t$ to be added to the bins at time step $t$. In our model, the ball $d_t$ associated with some request $\sigma_t$ is deleted at the beginning of time step $t$, next, the ball $a_t$ has to be placed into one of the $n$ bins, without any knowledge of the following requests $\sigma_{t+1}$, $\sigma_{t+2},\dots$ This infinite sequence of insertions and deletions is fixed in advance and is thus independent of previously allocated balls and any decisions made by the allocation algorithm. We assume that at any time $t$ at most $m = n$ balls are in the bins.

\section{Multiple-choice BiB}
\label{sec:multiple-choice}
The single-choice BiB strategy (Section \ref{sec:single-choice}) chooses one bin uniformly and independently at random for each ball. A multiple-choice BiB strategy generalizes this idea by  choosing $d \geq 2$ bins and placing the ball into the least full. We distinguish placement algorithms depending on how the $d$ locations (i.e, bins) are sampled~\cite{VOC03}.

\subsection{Classes of random placement strategies}
\label{sec:classesOfPlacement}
Let $B$ be the set of bins $\{1,2,...,n\}$ and $B^{d}$ be the sample space. Placement algorithms are distinguished by the way the $d$ candidate bins are chosen for each ball:
\begin{description}
\item [Uniform and independent:] Choose each of the $d$ bins for a ball uniformly and independently at random from the set of bins $B$. Thus, the probability that any specific bin $b$ is chosen for the $i$-th balls is
\begin{align*}
\Pr[\mathrm{bin}\;b\;is\;\mathrm{chosen}] = \frac{1}{n} \quad .
\end{align*}
\item [Non-uniform and independent:] Choose the $i$-th bin from $B$ independently at random, according to some probability distribution.
\item [Non-uniform and dependent:] Choose the $d$ bins for a ball from the sample space according to some probability probability distribution.
\end{description}

We will discuss two different multiple-choice placement strategies: The Greedy algorithm (Section \ref{sec:greedy}), belonging to class 1, and the Always-go-left scheme (Section \ref{sec:AlwaysGoLeft}), which belongs to class 2.

\subsection{The Greedy scheme}
\label{sec:greedy}
Consider the Greedy scheme: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the least full bin. If there are multiple bins with the same minimal load, chose one among them arbitrarily. Note that the $d$ selected bins are not necessarily distinct.

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, then the maximum load is at most
\begin{align*}
\load \leq \frac{\ln \ln n}{\ln d}+ O(1) \quad ,
\end{align*}
w.h.p., for all $t$~\cite{ABKU99, BCSV06}.
\end{theorem}
This result is remarkable in several ways:
\begin{compactitem}
\item The apparently small change made for the Greedy algorithm compared to the single-choice BiB scheme results in an exponential decrease of the gap, even for $d=2$. 
\item Each additional choice ($d > 2$) decreases the gap by only a constant factor~\cite{MRS01}.
\end{compactitem}

\subsection{The Always-go-left scheme}
\label{sec:AlwaysGoLeft}
The Greedy strategy presented in Section~\ref{sec:greedy} selects the $d$ bins uniformly and independently at random. Surprisingly, a multiple-choice BiB strategy with a non-uniform and independent selection of the bins, such as the Always-go-left algorithm introduced by V\"ocking in~\cite{VOC03}, achieves better load balancing than the Greedy scheme~\cite{BCSV06}.

The Always-go-left algorithm works as follows. We partition the $n$ bins into $d$ groups (each of size $n/d$), select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins and put the ball into the bin with smallest index among those with same minimal load.

\begin{theorem}
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load is at most 
\begin{align*}
\load \leq \frac{\ln\ln n}{d  \ln \Phi_d} + O(1) \quad ,
\end{align*}
w.h.p.,~for all $t$~\cite{VOC03}.
\end{theorem}

\section{Analysis}
\label{sec:analysis}
In this section, we will prove Theorems \ref{theorem:greedy} and \ref{theorem:aglm} in detail, using witness trees to provide an upper bound for the event that a bin contains too many balls. (Our analysis closely follows~\cite{VOC03}).

When the maximum load exceeds some given threshold we say that a bad event occurs and construct an activated witness tree. Thus, the activation of a witness tree is implied by a bad event. The probability for the existence of such an activated witness tree upper-bounds the probability that the associated bad event occurs.

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections~\ref{sec:analysis:greedy} and~\ref{sec:analysis:alg}, respectively. However, we will adopt a simplifying assumption to provide a more clear presentation of the proof technique. Section~\ref{sec:analysis:nondistinctBalls} will extend these proofs to remove this assumption.

We will use a proof technique called witness trees, which allows for a clean analysis of both problems and is applicable to a wide variety of BiB scenarios. Other proof techniques such as layered induction and a differential equation approaches have been used in the literature; see, e.g.,~\cite{MRS01} for an overview of different proof techniques for the BiB problem. The proof for arbitrarily large $m$ is more involved and we refer the interested reader to~\cite{BCSV06, TW13}.
